Analytic reproducibility in articles receiving open data badges at Psychological Science: An observational assessment

### 2020-11-17 manuscript/manuscript.Rmd
minor text changes (RSOS revisions round 2)

```{r load_packages}
# load packages
library(knitr) # for literate programming
library(papaja) # for article template
library(tidyverse) # for data munging
library(tidylog) # inline analysis feedback
library(here) # for finding files
library(DescTools) # for calculating confidence intervals
library(english) # for converting numbers to words
library(meta) # for meta-analysis
library(osfr) # to download data from the Open Science Framework (for meta-analysis)
```

@@

```{r load_functions}
# load functions for calculating confidence intervals and performing min-max normalization
source(here('analysis','functions.R'))
```

@@

A previous study in the domain of psychology largely circumvented data availability issues by capitalizing on a mandatory data sharing policy introduced at the journal Cognition [@hardwicke2018a]. In a sample of 35 articles with available data that had already passed initial quality checks, 24 (69%) articles contained at least one value that could not be reproduced within a 10% margin of error. Reproducibility issues were resolved in 11 of the 24 articles after consultation with original authors yielded additional data or clarified analytic procedures. Ultimately, 64 of 1324 (5%) checked values could not be reproduced despite author involvement. Importantly, there were no clear indications that the reproducibility issues seriously undermined the conclusions of the original articles. Nevertheless, this study highlighted a number of quality control and research transparency issues, including suboptimal data management; unclear, incomplete, or incorrect analysis specification; and reporting errors.

In the present study, we intended to investigate whether the findings of @hardwicke2018a extended to a corpus of Psychological Science articles that received an ‘open data badge’ to signal data availability. In order to focus specifically on the reproducibility of the analysis process rather than upstream data availability or data management practices, we selected only articles that had reusable data according to a previous study [@kidwell2016]. The submission guidelines of Psychological Science specifically stated that authors could earn an open data badge for “making publicly available the digitally-shareable data necessary to reproduce the reported result.” [^psychsci1] Additionally, authors were asked to self-certify that they had provided “…sufficient information for an independent researcher to reproduce the reported results.” [^psychsci2] Thus, if this policy was operating as intended, all numbers presented in these articles should be independently reproducible. If not, we hoped to learn about causes of non-reproducibility in order to identify areas for improvement in journal policy and research practice.

@@

We employed a retrospective non-comparative case-study design based on @hardwicke2018a. The numerical discrepancy between **original values** reported in the target articles and **reanalysis values** obtained in our reproducibility checks was quantified using percentage error[^percentError] ($PE = \frac{|reanalysis-original|}{original}\times100$). Numerical discrepancies were classified as ‘minor’ (0% > PE < 10%) or ‘major’ (PE >= 10%). If an old p-value fell on the opposite side of the $\alpha = .05$ boundary relative to a new p-value we additionally recorded a ‘decision error’. We recorded when there was insufficient information to complete aspects of the analyses.

@@

Of `r data_kidwell_badge %>% nrow()` articles marked with an open data badge, Kidwell and colleagues had identified `r data_kidwell_badge_reusable %>% nrow()` with datasets that met four reusability criteria (accessible, correct, complete, and understandable). For each of these articles, one investigator (TEH) attempted to identify a coherent set of descriptive and inferential statistics (e.g., means, standard deviations, t-values, p-values; see Figure \@ref(fig:valueTypePlot)), roughly 2-3 paragraphs of text, sometimes including a table or figure, related to a ‘substantive’ finding based on ‘relatively straightforward’ analyses. We focused on substantive findings because they are the most important and straightforward analyses to ensure that our team had sufficient expertise to re-run them. In total, `r nrow(data_values)` discrete numerical values reported in `r totalArticles` articles published between January 2014 and May 2015 were designated as target values. Further information about the sample is available in Supplementary Information \@ref(supsample).

@@

[^infStat]: Note that although one goal of the study was to characterize analytic reproducibility within a finite sample as a follow-up to Kidwell et al. (2016), we were also interested in generalizing to other psychology articles. To address this second goal, we use standard inferential statistics, recognizing that, because they are generated from a convenience sample, the generality of our findings may be limited.

@@

# prepare information on no author response
noResponse <- data_articles %>% filter(finalOutcome == 'Not reproducible\nwithout author\ninvolvement') %>% nrow()
```
Prior to seeking original author involvement, all target values in `r reproBefore %>% filter(reproOutcome == T) %>% pull(n)` out of the `r totalArticles` articles (`r reproBefore %>% filter(reproOutcome == T) %>% pull(percent)`) were reproducible, with the remaining `r reproBefore %>% filter(reproOutcome == F) %>% pull(n)` articles (`r reproBefore %>% filter(reproOutcome == F) %>% pull(percent)`) containing at least one major numerical discrepancy. After requesting input from original authors, several issues were resolved through the provision of additional information[^authorInfo] and ultimately all target values in `r reproAfter %>% filter(reproOutcome == T) %>% pull(n)` (`r reproAfter %>% filter(reproOutcome == T) %>% pull(percent)`) articles were reproducible, with the remaining `r reproAfter %>% filter(reproOutcome == F) %>% pull(n)` (`r reproAfter %>% filter(reproOutcome == F) %>% pull(percent)`) articles containing at least one major numerical discrepancy. In `r as.english(noResponse)` cases authors did not respond and in `r as.english(insuffInfoErrors)` cases there was insufficient information to proceed with the reanalysis. The full breakdown of reproducibility outcomes after author contact are shown in Table \@ref(tab:reproTab) and the results of a meta-analysis synthesizing the findings with those of a previous study [@hardwicke2018a] are displayed in Figure \@ref(fig:forestPlot). In no cases did the observed numerical discrepancies appear to be consequential for the conclusions stated in the original articles (see Supplementary Information \@ref(vignettes)).

[^authorInfo]: All except one case involved the provision of additional information that was not included in the original article or clarification of original information that was ambiguous or incomplete.In the exception case (4-1-2015_PS), the authors pointed out that we had missed a relevant footnote in the original article.

```{r reproTab}
data.frame("Target values fully reproducible?" = c("Yes", "Yes", "No", "No"),
           "Original author involvement?" = c("Yes", "No", "Yes", "No"),
           "n (%[95%CI])" = c(
             articleLevelSummary %>% 
               filter(finalOutcome == 'Reproducible\nwithout author\ninvolvement') %>%
               pull(fullReport),
             articleLevelSummary %>% 
               filter(finalOutcome == 'Reproducible\nwith author\ninvolvement') %>%
               pull(fullReport),
             articleLevelSummary %>% 
               filter(finalOutcome == 'Not reproducible\nwith author\ninvolvement') %>%
               pull(fullReport),
             articleLevelSummary %>% 
               filter(finalOutcome == 'Not reproducible\nwithout author\ninvolvement') %>%
               pull(fullReport))) %>% 
  kable(caption = "Final outcomes of reproducibility checks at the article level after original authors were contacted.", align = 'l', 
        col.names = c("Target values fully reproducible?",
                      "Original author involvement?",
                      "n (%[95%CI])"))
```

```{r forestPlot, fig.cap="Forest plot showing the proportion of articles with fully reproducible target values in Hardwicke et al. (2018) and the present study before and after contacting original authors for assistance with reproducibility checks. Squares represent individual study proportions (with size proportional to meta-analytic weighting). Diamonds represent summary proportions estimated by random effects (RE) models. Error bars represent 95% confidence intervals (CIs).", fig.width=8, fig.path="figs/"}
source(here('analysis','meta_analysis.R'))
```

@@

decisionErrors <- data_values %>% 
  filter(comparisonOutcome == "DECISION_ERROR") %>%
  nrow() # extract info on minor decision errors

@@

```{r scatterPlot, fig.cap="Scatterplot showing reanalysis p-values as a function of original p-values, classified by reproducibility outcome. Axes are on a log scale. The diagonal line represents perfect consistency between reanalysis and original values. Dashed lines represent the typical alpha threshold (0.05) for statistical significance. Original values reported relative to thresholds (e.g., p < .05) are represented by the threshold value when there was a major discrepancy and represented by the reanalysis value when there was a match. For display purposes, 41 values below 0.001 are not shown.", fig.path="figs/"}
source(here('analysis','plot_reproScatter.R')) # build plot
scatterPlot # show plot
```

After author involvement, `r majorErrors` major numerical discrepancies remained amongst the `r valuesChecked` target values examined across all articles (`r paste0(majorErrors_percent,'% ',majorErrors_percentCI)`). This included `r decisionErrors` decision error for which we obtained a statically significant p-value of `r data_values %>% filter(comparisonOutcome =='DECISION_ERROR') %>% pull(obtainedValue)` in contrast to the reported non-significant p-value of `r data_values %>% filter(comparisonOutcome =='DECISION_ERROR') %>% pull(reportedValue)`. A scatterplot illustrating the consistency of original and reanalysis p-values is displayed in Figure \@ref(fig:scatterPlot). The frequency of numerical discrepancies by value types is displayed in Figure \@ref(fig:valueTypePlot).

@@

@@

This study has a number of important limitations and caveats. Most pertinently, generalizing the observed reproducibility rates to other articles requires some caution as several characteristics of our sample are likely to have had a positive impact on the reproducibility rate. We specifically selected a sample of articles for which data were already available and screened against several reusability criteria [@kidwell2016]. By contrast, most psychology articles are unlikely to be accompanied by raw data or analysis scripts [@hardwicke2018; @hardwicke2020a; @wicherts2006], and even when data is available, suboptimal management and documentation can complicate re-use [@hardwicke2018a]. Consequently, most psychology articles would likely fail a reproducibility check before reaching the stage of the analysis pipeline at which we began our assessments.

@@

The study protocol (hypotheses, methods, and analysis plan) was pre-registered on the Open Science Framework on October 18, 2017 (https://osf.io/2cnkq/). All deviations from this protocol or additional exploratory analyses are explicitly acknowledged in Supplementary Information \@ref(prereg). We report how we determined our sample size, all data exclusions, all manipulations, and all measures in the study. All data, materials, and analysis scripts related to this study are publicly available on the Open Science Framework (https://osf.io/n3dej/). To facilitate reproducibility this manuscript was written by interleaving regular prose and analysis code using knitr [@R-knitr] and papaja [@R-papaja], and is available in a Code Ocean container (https://doi.org/10.24433/CO.1796004.v1) which re-creates the software environment in which the original analyses were performed. Analysis code, reports, and Code Ocean containers are also available for each reproducibility check (Supplementary Information \@ref(vignettes)).

>>>>

```{r load_packages}
# load packages
library(knitr) # for literate programming
library(papaja) # for article template
library(tidyverse) # for data munging
library(tidylog) # inline analysis feedback
library(here) # for finding files
library(DescTools) # for calculating confidence intervals
library(english) # for converting numbers to words
library(meta) # for meta-analysis
library(osfr) # to download data from the Open Science Framework (for meta-analysis)
```
```{r load_functions}
# load functions for calculating confidence intervals and performing min-max normalization
source(here('analysis','functions.R'))
```

@@

@@

A previous study in the domain of psychology largely circumvented data availability issues by capitalizing on a mandatory data sharing policy introduced at the journal Cognition [@hardwicke2018a]. In a sample of 35 articles with available data that had already passed initial quality checks, 24 (69%) articles contained at least one value that could not be reproduced within a 10% margin of error. Reproducibility issues were resolved in 11 of the 24 articles after consultation with original authors yielded additional data or clarified analytic procedures. Ultimately, 64 of 1324 (5%) checked values could not be reproduced despite author involvement. In some cases, the data files contained errors or were missing values and at least one author published a correction notice. Importantly, there were no clear indications that the reproducibility issues seriously undermined the conclusions of the original articles. Nevertheless, this study highlighted a number of quality control and research transparency issues including suboptimal data management; unclear, incomplete, or incorrect analysis specification; and reporting errors.

In the present study, we intended to investigate whether the findings of @hardwicke2018a extended to a corpus of Psychological Science articles that received an ‘open data badge’ to signal data availability. In order to focus specifically on the reproducibility of the analysis process rather than upstream data availability or data management practices, we selected only articles that had reusable data according to a previous study [@kidwell2016]. The submission guidelines of Psychological Science specifically stated that authors could earn an open data badge for “making publicly available the digitally-shareable data necessary to reproduce the reported result.” [^psychsci1] Additionally, authors were asked to self-certify that they had provided “…sufficient information for an independent researcher to reproduce the reported results.” [^psychsci2] If this policy was operating as intended, all numbers presented in these articles should be independently reproducible. If not, we hoped to learn about causes of non-reproducibility in order to identify areas for improvement in journal policy and research practice. Thus, the aim of the study was to assess the extent to which data shared under the Psychological Science open badge scheme actually enabled analytic reproducibility.

@@

We employed a retrospective non-comparative case-study design based on @hardwicke2018a. The numerical discrepancy between **original values** reported in the target articles and **reanalysis values** obtained in our reproducibility checks was quantified using percentage error[^percentError] ($PE = \frac{|reanalysis-original|}{original}\times100$). Numerical discrepancies were classified as ‘minor’ (0% > PE < 10%) or ‘major’ (PE >= 10%). If an old p-value fell on the opposite side of the alpha boundary relative to a new p-value we additionally recorded a ‘decision error’. The alpha boundary was assumed to be .05 unless otherwise stated. We recorded when there was insufficient information to complete aspects of the analyses.

@@

Of `r data_kidwell_badge %>% nrow()` articles marked with an open data badge, Kidwell and colleagues had identified `r data_kidwell_badge_reusable %>% nrow()` with datasets that met four reusability criteria (accessible, correct, complete, and understandable). For each of these articles, one investigator (T.E.H.) attempted to identify a coherent set of descriptive and inferential statistics (e.g., means, standard deviations, t-values, p-values; see Figure \@ref(fig:valueTypePlot)), roughly 2-3 paragraphs of text, sometimes including a table or figure, related to a ‘substantive’ finding based on ‘relatively straightforward’ analyses. We focused on substantive findings because they are the most important and straightforward analyses to ensure that our team had sufficient expertise to re-run them. In total, `r nrow(data_values)` discrete numerical values reported in `r totalArticles` articles published between January 2014 and May 2015 were designated as target values. Further information about the sample is available in Supplementary Information \@ref(supsample).

@@

[^infStat]: Note that although one goal of the study was to characterize analytic reproducibility within a finite sample as a follow-up to Kidwell et al. (2016), we were also interested in generalizing to other psychology articles. To address this second goal, we use standard inferential statistics, recognizing that, because they are generated from a convenience sample, the generality of our findings may be limited (for more details, see discussion).

@@

# prepare information on no author response
noResponse <- data_articles %>% filter(finalOutcome == 'Not reproducible\nwithout author\ninvolvement') %>% nrow()
```
Prior to seeking original author involvement, all target values in `r reproBefore %>% filter(reproOutcome == T) %>% pull(n)` out of the `r totalArticles` articles (`r reproBefore %>% filter(reproOutcome == T) %>% pull(percent)`) were reproducible, with the remaining `r reproBefore %>% filter(reproOutcome == F) %>% pull(n)` articles (`r reproBefore %>% filter(reproOutcome == F) %>% pull(percent)`) containing at least one major numerical discrepancy. After requesting input from original authors, several issues were resolved through the provision of additional information and ultimately all target values in `r reproAfter %>% filter(reproOutcome == T) %>% pull(n)` (`r reproAfter %>% filter(reproOutcome == T) %>% pull(percent)`) articles were reproducible, with the remaining `r reproAfter %>% filter(reproOutcome == F) %>% pull(n)` (`r reproAfter %>% filter(reproOutcome == F) %>% pull(percent)`) articles containing at least one major numerical discrepancy. In all except one case author input involved provision of additional information that was not included in the original article or clarification of original information that was ambiguous or incomplete. In the exception case (4-1-2015_PS), the authors pointed out that we had missed a relevant footnote in the original article. In `r as.english(noResponse)` cases authors did not respond and in `r as.english(insuffInfoErrors)` cases there was insufficient information to proceed with the reanalysis. The full breakdown of reproducibility outcomes after author contact are shown in Table \@ref(tab:reproTab) and the results of a meta-analysis synthesizing the findings with those of a previous study [@hardwicke2018a] are displayed in Figure \@ref(fig:forestPlot). In no cases did the observed numerical discrepancies appear to be consequential for the conclusions stated in the original articles (see Supplementary Information \@ref(vignettes)).

```{r reproTab}
data.frame("Target values fully reproducible?" = c("Yes", "Yes", "No", "No"),
           "Original author involvement?" = c("Yes", "No", "Yes", "No"),
           "n (%[95%CI])" = c(
             articleLevelSummary %>% 
               filter(finalOutcome == 'Reproducible\nwithout author\ninvolvement') %>%
               pull(fullReport),
             articleLevelSummary %>% 
               filter(finalOutcome == 'Reproducible\nwith author\ninvolvement') %>%
               pull(fullReport),
             articleLevelSummary %>% 
               filter(finalOutcome == 'Not reproducible\nwith author\ninvolvement') %>%
               pull(fullReport),
             articleLevelSummary %>% 
               filter(finalOutcome == 'Not reproducible\nwithout author\ninvolvement') %>%
               pull(fullReport))) %>% 
  kable(caption = "Final outcomes of reproducibility checks at the article level after original authors were contacted. Note that although one confidence interval technically includes 0, we are confident that 0 is not a possible population proportion due to the existence of non-reproducible cases that we document here.", align = 'l', 
        col.names = c("Target values fully reproducible?",
                      "Original author involvement?",
                      "n (%[95%CI])"))
```

```{r forestPlot, fig.cap="Forest plot showing the proportion of articles with fully reproducible target values in Hardwicke et al. (2018) and the present study before and after contacting original authors for assistance with reproducibility checks. Squares represent individual study proportions. Diamonds represent summary proportions estimated by random effects (RE) models. Error bars represent 95% confidence intervals (CIs).", fig.width=8, fig.path="figs/"}
source(here('analysis','meta_analysis.R'))
```

@@

decisionErrors <- data_values %>% 
  filter(comparisonOutcome == "DECISION_ERROR") %>%
  nrow() # extract info on decision errors

decisionErrors_sig <- data_values %>% 
  filter(comparisonOutcome == "DECISION_ERROR",
         reportedValue >= .05) %>%
  nrow() # extract info on decision errors where reported value was statistically significant

decisionErrors_nonsig <- data_values %>% 
  filter(comparisonOutcome == "DECISION_ERROR",
         reportedValue < .05) %>%
  nrow() # extract info on decision errors where reported value was not statistically significant

@@

```{r scatterPlot, fig.cap="Scatterplot showing reanalysis p-values as a function of original p-values, classified by reproducibility outcome. Axes are on a log scale. The diagonal line represents perfect consistency between reanalysis and original values. Dashed lines represent the typical alpha threshold (0.05) for statistical significance. Original values reported relative to thresholds (e.g., p < .05) are represented by the threshold value when there was a major discrepancy and represented by the reanalysis value when there was a match. For display purposes, 41 values below 0.001 are not shown. Note that the two values in the bottom right quadrant of the graph have been conservatively designated as ‘major errors’ rather than ‘decision errors’ due to uncertainty about how the original analysis was performed. Specifically, insufficient information was provided in the original article (ID 9-5-2014_PS) about how multiplicity corrections were applied (see Vignette 25 in Supplementary Information E for further information).", fig.path="figs/"}
source(here('analysis','plot_reproScatter.R')) # build plot
scatterPlot # show plot
```

After author involvement, `r majorErrors` major numerical discrepancies remained amongst the `r valuesChecked` target values examined across all articles (`r paste0(majorErrors_percent,'% ',majorErrors_percentCI)`). This included `r decisionErrors_sig` decision errors for which we obtained a statically significant p-value in contrast to a reported non-significant p-value and `r decisionErrors_nonsig` decision error with the opposite pattern. A scatterplot illustrating the consistency of original and reanalysis p-values is displayed in Figure \@ref(fig:scatterPlot). The frequency of numerical discrepancies by value types is displayed in Figure \@ref(fig:valueTypePlot).

@@

Non-reproducibility does not always imply that prior conclusions based on the original results are fatally undermined – indeed we encountered no such scenarios in this study. Determining the consequences of non-reproducibility for the interpretation of the results is not always straightforward and is to some extent subjective. We considered several factors including the precision of original hypotheses, the extent and magnitude of non-reproducibility, whether there were p-value decision errors. For example, in a case where a hypothesis only makes a directional prediction, a reanalysis indicating a statistically significant but smaller effect than originally reported, would still be consistent with the original hypothesis. Or in a case where effect sizes and p-values can be reproduced successfully, but one finds a few major numerical errors in the descriptive statistics, the most obvious cause is a reporting error, rather than an error in the implementation of the original analysis, and the original conclusions are not obviously compromised by this. Therefore, scientific reasoning is needed to evaluate the consequences of non-reproducibility on a case-by-case basis.

@@

This study has a number of important limitations and caveats. Most pertinently, we have reported confidence intervals and conducted a random-effects meta-analysis to aid generalization of the findings to a broader population; however, such generalizations should be made with caution as several characteristics of our sample are likely to have had a positive impact on the reproducibility rate. We specifically selected a sample of articles for which data were already available and screened against several reusability criteria [@kidwell2016]. By contrast, most psychology articles are unlikely to be accompanied by raw data or analysis scripts [@hardwicke2018; @hardwicke2020a; @wicherts2006], and even when data is available, suboptimal management and documentation can complicate re-use [@hardwicke2018a]. Consequently, most psychology articles would likely fail a reproducibility check before reaching the stage of the analysis pipeline at which we began our assessments.

@@

The study protocol (hypotheses, methods, and analysis plan) was pre-registered on the Open Science Framework on October 18, 2017 (https://osf.io/2cnkq/). All deviations from this protocol or additional exploratory analyses are explicitly acknowledged in Supplementary Information \@ref(prereg). We report how we determined our sample size, all data exclusions, all manipulations, and all measures in the study. All data, materials, and analysis scripts related to this study are publicly available on the Open Science Framework (https://osf.io/n3dej/). To facilitate reproducibility this manuscript was written by interleaving regular prose and analysis code using knitr [@R-knitr] and papaja [@R-papaja], and is available in a Code Ocean container (https://doi.org/10.24433/CO.1796004.v2) which re-creates the software environment in which the original analyses were performed. Analysis code, reports, and Code Ocean containers are also available for each reproducibility check (Supplementary Information \@ref(vignettes)).
